---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# greenclust

## Overview

The `greenclust` package implements a method of grouping/clustering the categories of a contingency table in a way that preserves as much of the original variance as possible. It is well-suited for reducing the number of levels of a categorical feature in logistic regression (or any other model having a categorical outcome), while still maintaining some degree of explanatory power.

It does this by iteratively collapsing the rows two at a time, similar to other agglomerative hierarchical clustering methods. At each step, it selects the pair of rows whose combination results in a new table with the smallest loss of chi-squared. The process is often refered to "Greenacre's Method", particularly in the SAS community, after statistician Michael J. Greenace.

The returned object is an extended version of the `hclust` object used in the `stats` package and can be used similarly (plotted as a dendrogram, cut, etc.). Additional functions are provided in the package for automatic cutting and diagnostic plotting.

## Installation

You can install the current version of `greenclust` directly from this GitHub repository using the [devtools package](https://www.rstudio.com/products/rpackages/devtools/):

``` r
# install.packages("devtools")
devtools::install_github("jeffjetton/greenclust")
```

## Example

The `greenclust()` function works like `hclust()`, only it accepts a contingency table rather than a dissimilarity matrix. For the purposes of this example, we'll merge the categorical features of the [Titanic](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/Titanic.html) data set into a single, monolithic category. 

```{r make.table}
# Combine Titanic passenger attributes into a single category
tab <- t(as.data.frame(apply(Titanic, 4:1, FUN=sum)))

# Remove rows with all zeros (not valid for chi-squared test)
tab <- tab[apply(tab, 1, sum) > 0, ]
```

This gives us a contingency table with several levels, showing the total number of passengers who survived (or not) at each level:

```{r show.table, echo=FALSE}
tab
```

From there, we can perform our clustering:

```{r example_clusterplot_1}
# Create greenclust tree object from table
library(greenclust)
grc <- greenclust(tab)

# This can be plotted like any standard hclust tree
plot(grc)
```

The "height" in this case is the reduction in r-squared. That is, the proportion of chi-squared, relative to the original uncollapsed table, that is lost at each given clustering node.

The package provides a special plotting function, `greenplot(), that shows the r-squared and chi-squared test p-value for each potential number of groups/clusters. This can be a useful aid in weighing the trade-off between fewer clusters and lower r-squared:

```{r example_greenplot}
greenplot(grc)
```

When using this method, the customary "optimal" number of groups is found at most-significant chi-squared test (i.e., lowest p-value). This point is automatically highlighted on the above chart.

The `greencut()` function is essentially a version of `cutree()` that automatically cuts a greenclust tree at this optimal level:

```{r greencut}
greencut(grc)
```

Note that `greencut()` also includes the r-squared and p-value for that particular clustering level as vector attributes. If you prefer a different cut point, but still would like to have these attributes, you can override automatic selection by specifying either `k` (number of clusters) or `h` (height, or 1 - r-squared):

```{r greencut2}
greencut(grc, k=3)
```
   
---
   
Putting it all together:

```{r example_clusterplot2}
clusters <- greencut(grc)

plot(grc)
rect.hclust(grc, max(clusters), border=unique(clusters)+1)

new.tab <- aggregate(tab, by=list(clusters), FUN=sum)[ , 2:3]
row.names(new.tab) <- paste0("clust", unique(clusters))
new.tab

# New table still shows clear differences in survival rates
round(prop.table(as.matrix(new.tab), 1), 3)
```
